---
title: "Non-opioid analgesics reg sci project_scratch"
author: "Cat Luria"
date: "3/29/2021"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(ggplot2)
library(stringr)
library(forcats)
library(dplyr)
library(survival)
library(ggfortify)
library(tidyverse)
library(survminer)


```

### Defining a cohort

```{r echo=FALSE, message=FALSE}

#drugs with WHO ATC N02, dowloaded from Adis Insights on 2021-03-29 (all indications)
drugs_ATC_NO2_expanded <- read.delim("../data/drugs.txt", stringsAsFactors=FALSE)
drugs_ATC_NO2_expanded_dt <- data.table(drugs_ATC_NO2_expanded)

```


```{r echo=FALSE, message=FALSE}

## add column Highest.Development.Phase.Simpler which only states what phase drug is in regardless of status and column Highest.Development.Phase.Activity that states whether development is active or inactive. "discontinued", "suspended", "no development reported", "withdrawl" all categorized as inactive.

atc_n02_phases <- drugs_ATC_NO2_expanded_dt[, Highest.Development.Phase.Simple := Highest.Development.Phase]
atc_n02_phases$Highest.Development.Phase.Activity <- "Active"


##remove rows with uninformative reported highest development phase
atc_n02_phases <- atc_n02_phases[ ! Highest.Development.Phase.Simple %like% "Unknown"]
atc_n02_phases <- atc_n02_phases[ ! Highest.Development.Phase.Simple %like% "\\(Clinical\\)"]
atc_n02_phases <- atc_n02_phases[ ! Highest.Development.Phase.Simple %like% "0"]
atc_n02_phases <- atc_n02_phases[ Highest.Development.Phase.Simple != "Discontinued"]
atc_n02_phases <- atc_n02_phases[ Highest.Development.Phase.Simple != "No development reported"]
################################


##Categorize all statuses "No development reported", "Suspended", "Discontinued", "Withdrawal" as "Inactive" 
length(atc_n02_phases$Highest.Development.Phase
       [atc_n02_phases$Highest.Development.Phase %like% "No development reported"]) +
  length(atc_n02_phases$Highest.Development.Phase
         [atc_n02_phases$Highest.Development.Phase %like% "Discontinued" ]) +
  length(atc_n02_phases$Highest.Development.Phase
         [atc_n02_phases$Highest.Development.Phase %like% "Suspended" ]) +
length(atc_n02_phases$Highest.Development.Phase
         [atc_n02_phases$Highest.Development.Phase %like% "Withdrawal" ]) 


atc_n02_phases$Highest.Development.Phase.Activity[atc_n02_phases$Highest.Development.Phase %like% "No development reported" ] <- "Inactive"
atc_n02_phases$Highest.Development.Phase.Activity[atc_n02_phases$Highest.Development.Phase %like% "Discontinued" ] <- "Inactive"
atc_n02_phases$Highest.Development.Phase.Activity[atc_n02_phases$Highest.Development.Phase %like% "Suspended" ] <- "Inactive"
atc_n02_phases$Highest.Development.Phase.Activity[atc_n02_phases$Highest.Development.Phase %like% "Withdrawal" ] <- "Inactive"

length(atc_n02_phases$Highest.Development.Phase.Activity
         [atc_n02_phases$Highest.Development.Phase.Activity == "Inactive" ]) 

################################


## Change all statuses to simplest version: Research, Preclinical Phases I-III, Preregistration, Registered, and Marketed
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(Research\\)" ] <- "Research"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(Preclinical\\)" ] <- "Preclinical"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(I\\)" ] <- "Phase I"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "Phase I/II" ] <- "Phase I"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(I/II\\)" ] <- "Phase I"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "Phase II/III" ] <- "Phase II"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(II\\)" ] <- "Phase II"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(II/III\\)" ] <- "Phase II"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(III\\)" ] <- "Phase III"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(Registered\\)" ] <- "Registered"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "\\(Preregistration\\)" ] <- "Preregistration"
atc_n02_phases$Highest.Development.Phase.Simple[atc_n02_phases$Highest.Development.Phase.Simple %like% "Market" ] <- "Marketed"

length(atc_n02_phases$Highest.Development.Phase[atc_n02_phases$Highest.Development.Phase.Simple=="Marketed" &
                                                         atc_n02_phases$Highest.Development.Phase.Activity == "Inactive"])


```


```{r echo=FALSE, message=FALSE}


## add new column Highest.Development.Phase.Survival and convert phase from Highest.Development.Phase.Simple into a numeric score in order to generate "survival curve"

atc_n02_phases <- atc_n02_phases[, Highest.Development.Phase.Survival := Highest.Development.Phase.Simple]
atc_n02_phases <- atc_n02_phases[, Highest.Development.Phase.Status := Highest.Development.Phase.Activity]


atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Research"] <- 1
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Preclinical"] <- 2
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Phase I"] <- 3
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Phase II"] <- 4
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Phase III"] <- 5
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Preregistration"] <- 6
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Registered"] <- 7
atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == "Marketed"] <- 8
 


atc_n02_phases$Highest.Development.Phase.Status[atc_n02_phases$Highest.Development.Phase.Activity == "Active"] <- 0
atc_n02_phases$Highest.Development.Phase.Status[atc_n02_phases$Highest.Development.Phase.Activity == "Inactive"] <- 1     

a <- atc_n02_phases$Highest.Development.Phase.Survival[atc_n02_phases$Highest.Development.Phase.Survival == 8]
length(a)

  atc_n02_phases %>%
  group_by(Highest.Development.Phase.Status, Highest.Development.Phase.Survival) %>%
  summarise(
    n = n(),
  )

```



```{r echo=FALSE, message=FALSE}

#atc_n02_phases <- atc_n02_phases[,atc_n02_phases$ATC]

N02A_cols <- colnames(atc_n02_phases %>% select(contains("N02A")))
N02B_cols <- colnames(atc_n02_phases %>% select(contains("N02B")))

atc_n02_phases$ATC <- "other"

atc_n02_phases$ATC[rowSums(data.frame(atc_n02_phases)[,N02A_cols]) > 0 &
                     rowSums(data.frame(atc_n02_phases)[,N02B_cols]) == 0] <- "N02A"

atc_n02_phases$ATC[rowSums(data.frame(atc_n02_phases)[,N02B_cols]) > 0 &
                     rowSums(data.frame(atc_n02_phases)[,N02A_cols]) == 0] <- "N02B"

atc_n02_phases <- atc_n02_phases[atc_n02_phases$ATC!="other",]


atc_n02_phases$Highest.Development.Phase.Survival <- as.numeric(atc_n02_phases$Highest.Development.Phase.Survival)

atc_n02_phases$Highest.Development.Phase.Status <- as.numeric(atc_n02_phases$Highest.Development.Phase.Status)


atc_n02_phases_plot <- 
  atc_n02_phases %>%
  group_by(ATC, Highest.Development.Phase.Status, Highest.Development.Phase.Survival) %>%
  summarise(
    n = n(),
  )

p <- atc_n02_phases %>%
  group_by(ATC, Highest.Development.Phase.Status) %>%
  summarise(
    n = n(),
  )

```


All drugs in WHO ATC class N02 were downloaded from Adis Insight database on March 29, 2021. Of these, 329 were opioids (N02A) and 427 were non-opioids (N02B). Drugs classified as both N02A and N02B were excluded, leaving a total cohort of 756 drugs. Drugs with latest phase of development indicating suspension, discontinuation, no development reported for several years or market withdrawl were categorized here as "inactive". Similar proportions--approximately one third--of both N02A and N02B drugs were "active". 


```{r echo=FALSE}

#remove drugs that have not been significantly updated in the last 10 years

#atc_n02_phases2$Last.Significant.Update.Date.Date <- as.Date(atc_n02_phases2$Last.Significant.Update.Date, "%d/%m/%Y")
#atc_n02_phases3 <- atc_n02_phases2[atc_n02_phases2$Last.Significant.Update.Date.Date>="2011-05-01",]

atc_n02_phases %>%
  group_by(ATC) %>%
  summarise(
    n = n(),
  )

```

```{r echo=FALSE}

#generate list of drugs that are Phase 1 or later to use in searching ct.gov

# atc_n02_phases_export <- atc_n02_phases3[atc_n02_phases3$Highest.Development.Phase.Survival>=2,] 
# drug_list <- atc_n02_phases_export$Drug.Name
# write.csv(rbind(drug_list), file="data/drug_list_20210503.csv", row.names=FALSE)

```


```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of opioid (N02A) and non-opioid (N02B) analgesics in active development and inactive (e.g. 'suspended', 'discontinued', 'no development reported', or 'withdrawn') among phases of development. Drugs with no significant updates within last ten years are not shown."}

phases <- c(
  "1" = "Research",
  "2" = "Preclin",
  "3" = "Phase I",
  "4" = "Phase II",
  "5" = "Phase III",
  "6" = "Prereg",
  "7" = "Registered",
  "8" = "Marketed"
)

atc_n02_phases_plot <- 
  atc_n02_phases %>%
  group_by(ATC, Highest.Development.Phase.Status, Highest.Development.Phase.Survival) %>%
  summarise(
    n = n(),
  )


classes <- c("Non-opioid", "Opioid")

ggplot(atc_n02_phases_plot,
       aes(fill=as.factor(Highest.Development.Phase.Status), x=ATC, y=n)) +
    geom_bar(position = position_stack(reverse = TRUE), stat="identity") +
  scale_x_discrete(labels = classes) +
 scale_fill_discrete(labels=c("Active", "Inactive"), name="Status") +
  labs(x="Drug class") +
  facet_grid(Highest.Development.Phase.Survival~., labeller=labeller(Highest.Development.Phase.Survival=phases)) +
  coord_flip() +
  theme(legend.position="bottom", legend.direction = "horizontal", strip.text.y = element_text(angle=0))

```

```{r echo=FALSE, fig.cap="Survival curve showing at which stage of development opioid (N02A) and non-opioid (N02B) analgesics shift from active development to inactive (e.g. 'suspended', 'discontinued', 'no development reported', or 'withdrawn'), only drugs significantly updated in the last decade."}

fit <- survfit(Surv(atc_n02_phases$Highest.Development.Phase.Survival, atc_n02_phases$Highest.Development.Phase.Status)~ATC, data = atc_n02_phases)
autoplot(fit, conf.int = TRUE)

survdiff(Surv(atc_n02_phases$Highest.Development.Phase.Survival, atc_n02_phases$Highest.Development.Phase.Status)~ATC, data = atc_n02_phases)

#No significant difference

# ggsurvplot(fit, xlab="Development phase", ylab="Porportion of drugs that entered phase") + scale_x_discrete(limits=c("Research", "Preclin", "Phase 1", "Phase 2", "Phase 3", "Prereg", "Registered", "Marketed"))
# 
# x <- c(1,fit$time)
# y <- c(1,(fit$surv))
# plot(x,y, type="S")
# 
# fit$surv
# 
# strata <- summary(fit)$strata
# time <- summary(fit)$time
# surv <- summary(fit)$surv
# 
# surv_output <- as.data.table(strata)
# surv_output$time <- time
# surv_output$surv <- surv
# 
# ggplot(data=surv_output, aes(x=time, y=surv, color=strata)) + geom_line()

```

```{r echo=FALSE}

#atc_n02_phases_export <- atc_n02_phases3[, c(1:37, 1919:1923),]
# drug_list <- atc_n02_phases_export$Drug.Name
#write.csv(atc_n02_phases_export, file="data/filtered_drugs_20210505.csv", row.names=TRUE)

```

```{r echo=FALSE}

#check if biomarkers = activity

atc_n02_phases$Biomarkers.Logic <- apply(atc_n02_phases[,32:36],MARGIN = 1, function(x) sum(x!=""))>0

chi.table <- table(atc_n02_phases$Highest.Development.Phase.Activity, atc_n02_phases$Biomarkers.Logic)
chisq.test(chi.table)

chi.table[2,2]/(chi.table[1,2]+chi.table[2,2])
chi.table[2,1]/(chi.table[1,1]+chi.table[2,1])

#check if there is a significant difference between opioids and non-opioids in terms of what fraction have reported biomarkers

chi.table <- table(atc_n02_phases$Biomarkers.Logic, atc_n02_phases$ATC)
chi.table
chisq.test(chi.table)

chi.table[2,2]/(chi.table[1,2]+chi.table[2,2])
chi.table[2,1]/(chi.table[1,1]+chi.table[2,1])

#Check to see if there is a difference between development phases in fractions of drugs that have biomarkers

kt <- kruskal.test(x=atc_n02_phases$Highest.Development.Phase.Survival, g=atc_n02_phases$Biomarkers.Logic)

```


```{r}

### barplot of top indications

at <- atc_n02_phases[,grepl("IND", colnames(atc_n02_phases)), with=FALSE]

sumdata=data.frame(value=apply(at,2,sum))
sumdata$key=colnames(at)
sumdata <- sumdata[sumdata$value!=0,]
sumdata <- sumdata[order(-sumdata$value),]

sumdata$key <- factor(sumdata$key, levels = sumdata$key)

ggplot(data=sumdata[1:10,], aes(x=key, y=value, fill=key)) +
geom_bar(colour="black", stat="identity")




```


```{r}

# at <- atc_n02_phases[atc_n02_phases$ATC=="N02A",grepl("MCH", colnames(atc_n02_phases)), with=FALSE]
# 
# sumdata=data.frame(value=apply(at,2,sum))
# sumdata$key=colnames(at)
# sumdata <- sumdata[sumdata$value!=0,]
# sumdata <- sumdata[order(-sumdata$value),]
# 
# sumdata$key <- factor(sumdata$key, levels = sumdata$key)
# 
# ggplot(data=sumdata[1:10,], aes(x=key, y=value, fill=key)) +
# geom_bar(colour="black", stat="identity")
```

```{r echo=FALSE, fig.cap="Top 10 reported mechanisms of actions for N02B drugs."}
at <- atc_n02_phases[atc_n02_phases$ATC=="N02B",grepl("MCH", colnames(atc_n02_phases)), with=FALSE]

sumdata=data.frame(value=apply(at,2,sum))
sumdata$key=colnames(at)
sumdata <- sumdata[sumdata$value!=0,]
sumdata <- sumdata[order(-sumdata$value),]

sumdata$key <- factor(sumdata$key, levels = sumdata$key)

ggplot(data=sumdata[1:10,], aes(x=key, y=value, fill=key)) +
geom_bar(colour="black", stat="identity") +
  ylab("No. of drugs with MOA") +
 scale_fill_discrete(name = "MOA", labels=c("Undefined" ,"COX inhibitor", "COX2 inhibitor", "NMDA receptor antagonist", "Na channel antagonist", "CACNA2D1 modulator", "NK-1 antagonist", "Ca channel antagonist", "Cannabinoid receptor antagonist", "Glutamate receptor antagonist"))+
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```